Composition-climate paper

### 2015-11-18 Composition_Ecology.Rmd
Changes to main Rmd: * trying to clear up abstract conflict between bullets & text. * Changed the version of the PLS we're pulling from. * Changed the way the figures are pulled in the text.  Stopped making them in text and now pulling them in from file. * Some renaming of references for proper rendering. * Moved Figure 2 further down in the text. * Trying to clarify the language around the comparisons. * Fixed up Table 1.

Abstract
====
* Much of our understanding of species-climate relationships and functional-community relationships in forests is based on contemporary observations of tree distributions.
* Ecosystems are highly dynamic and contemporary forest have a heavy imprint of land use, while climate changes over decadal and centennial scales.
* Given long time-scale climate change and continued land use conversion, are species-climate relationships today representative of those in past.
* We assess the stability of species-climate relationships using historic PLS data and FIA data.
* We show that species-climate relationships have significantly shifted for many taxa.
* Contemporary trees have shifted to cooler maximum temperatures, warmer winter temperatures, reduced seasonal temperature range, and wetter conditions.
* Relative importance of land use and climate change varies by taxon and climate variable.  
* Shifts in warmer temperature mainly due to land use change, etc.
* Implications sentence or two at the end.

@@

@@

pls_data <- get_pls('0.9-4')
agg_dens <- get_fia()[,colnames(pls_data)]

@@

Over the past 200 years centennial scale climate variability and land use change have interacted to transform the forests of the northeastern United States. There are clear indications that modern forests, as represented by the Forest and Inventory Analysis show greater homogeneity and significantly different species associations than pre-settlement forests [@goring2015preset;@schulte2007homogenization]. This shift has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change [@iverson2013tree;@iverson1998predicting].  One of the most significant changes to forest tree distributions in the upper midwest is the loss of extensive cover through land use change [@rhemtulla2009legacies], particularly along the historical prairie and savanna margins of the region (Figure 1f), where extensive land use conversion has left little natural land.

@@

grid.arrange(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4))

@@

Methods
===============
We pair pre-settlement vegetation data from the Public Land Survey System (PLSS) aggregated to a 64 $km^{2}$ grid [@goring2015preset] with annually resolved climate data for the region.  The PLSS era data is aggregated from survey data collected on a 1 x 1 mi grid across the Upper Midwest during the PLS, from 1830 - 1904, in a time transgressive manner, from the south east to the north west.  At each survey point surveyors noted the closest two to four trees, recorded their distances to the plot center point, diameters and the azimuth to the tree.  Goring et al. [-@goring2015preset] aggregated this data to an 8 x 8 km grid to reduce local-scale variability, resulting in a gridded data product with an average of 61 plots per cell, or approximately 120 trees per cell.  When examining relationships between FIA and PLS data, Goring et al. [-@goring2015preset] report significant relationships within the FIA relating dissimilarity from PLSS era forests and the number of FIA plots within an 8x8km cell, however this relationship accounts for only 3% of the total change in dissimilarity, indicating statistical significance, but little ecological significance on this regional scale.

@@

The analysis presented here follows a blocked sampling method (Figure 2). We generate probability density functions along climate gradients for actual PLSS presence, by taxon, at the settlement era and for the same PLS taxa (and point locations) in the modern era. We do the same for FIA presence, and for anthropogenic land use classes. Climate data is derived from monthly PRISM data from 1895 - 2014. A 30-year climate was generated for modern July temperature, January temperature, the continentality (difference between T_jul_ and T_jan_), and annual precipitation.

![model grid](figures/model_grid.png)

**Figure 2**. *Blocked method for analysis.  Data presented represent one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row).  A diagonal shift from the top left to bottom right represents the true change in forest-climate relationships.*

@@

By generating kernel density estimates of taxon presence along climate gradients drawn from each of the four states of climate and vegetation (Figure 2) we can use Hellinger distance to estimate the effects of changing climate (d_H_ between kernel densities in the same column of Figure 2), or the effect of changing land use (d_H_ between kernel densities in the same row of Figure 2).  In particular, the effect of changing land use and climate (d_H_ between the top left and bottom right in Figure 2) can be parsed as the sum of differences in the left hand column (testing climate change) and the top row (testing land use only).  The relative magnitudes of these differences gives us an indication of the influence of either land use or climate in driving shifts in species distributions along the climate gradients in the Upper Midwest.

@@

loss_table <- as.data.frame.matrix(table(loss_melt[[1]]$variable,
                                         loss_melt[[1]]$value))
loss_table$Gain <- round(loss_table$Gain / (loss_table$Loss + loss_table$Presence)*100,1)

loss_table[,2:3] <- round(tran(loss_table[,2:3], method = 'proportion') * 100,1)

loss_melt[[2]]
```

**Figure 3.**.  *Gain loss and continuous presence for PLSS and FIA tree taxa. Loss is assessed where previously PLSS data indicated presence of a taxon but FIA data do not record presence (light blue). Continuous presence is indicated by dark gray, for where both FIA and PLSS cells indicate presence.  For most taxa it is the central parts of their distribution that show continuous presence.  Few taxa show novel presence (red), although taxa such as poplar, ash and basswood show gains of over 10%.*

**Table 1.** *Gain and loss of the various forest tree types of interest since the pre-settlement era. Gain since the pre-settlement era is the number of new cells as a proportion of historical cover for the tree taxon, thus positive values are not indicative of higher overall spatial coverage. Loss and Presence columns sum to 100% and represent the proprtion of the historical range that has been lost, or shown continuous presence.*

@@

```{r densities, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}

source('R/resampling_climate.R')
vegclim_table <- resample_climate(climate, agg_dens, pls_data)

source('R/figures/climate_boxplots.R')
boxplot_clim <- clim_boxplots(clim_table = vegclim_table)
boxplot_clim

@@

**Figure 4.** *Changes in the realized climate space for genera in the upper Midwestern United States. For forest taxa maximum temperatures decline across the region from the PLS era to the modern era.  At the same time minimum temperatures and annual precipitation have both increased.*

@@

```{r climate_shifts, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}

@@

hell_plot[[1]]

@@

**Figure 5**. *For each of the four variables examined we see differences in the relative contributions of land use and climate in driving patterns of change over the settlement era.  Taxa with negative y values have changes in distributions along the climate gradient that are driven largely by land use, while positive y axis values can be largely attributed to regional climate change.*

The Hellinger distances reveal that while taxa may show the imprint of land use change on shifting climatic niches, this shift is not uniform across taxa not climatic  variables. Maximum temperature shows the greatest change attributed to land use change.  Tamarack, elm, poplar, spruce beech, cedar, and fir all show a greater influence for land use change than regional climate change.

The precipitation panel (Figure 5) indicates that precipitation across the region has driven change in the species climate niche for this variable.  Only spruce and poplar show a marginally greater effect of land use change than climate change, wheras taxa such as beech and hemlock show a strong climate signal.

@@

Land use is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000]. Land use change reduces the correlational structure between species and climate [@devineau2011arid]. While rates of land use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land use are likely to persist, both in reorganizing forested landscapes [@goring2015], and in the correlational structure between species and climate. Paleoclimatic modeling indicates shifts in realized niches for species over time, potentially resulting from non-analogue climates or shifts in competition resulting from changing patterns of co-occurence through time [@maiorano2013niche].

From a modern perspective, the US Forest Service's Forest Inventory and Analysis data set is an important resource for researchers, providing detailed data at fine grain over the entire contiguous United States. However, the FIA is also an incomplete record of tree species distributions in the United States since it limits plot surveys to plots with > 10% tree cover [@woudenberg2010forest].  In much the same way, climatic data plays an important role in building correlative models, but we are limited to the modern era [@pellatt2012using;@iverson1998predicting], while evidence strongly suggests historical datasets add considerably to our understanding of the breadth of species' climatic niches [@ivory2015; @maiorano2013niche].

>>>>

Abstract
====
* Ecosystems are highly dynamic and contemporary forest have a heavy imprint of land use, while climate changes over decadal and centennial scales.
* Given long time-scale climate change and continued land use conversion, are species-climate relationships today representative of those in past.
* We assess the stability of species-climate relationships using historic PLS data and FIA data.
* We show that species-climate relationships have significantly shifted for many taxa.
* Contemporary trees have shifted to cooler maximum temperatures, warmer winter temperatures, reduced seasonal temperature range, and wetter conditions.
* Relative importance of land use and climate change varies by taxon and climate variable.  
* Shifts in warmer temperature mainly due to land use change, etc.
* Implications sentence or two at the end.

@@

*Using contemporary and historical species distributions and climate data we show that modern forest tree genera occupy realized niches with cooler maximum temperatures, higher minimum temperatures and higher rainfall than they have in the past.*

@@

pls_data <- get_pls('0.9-6')
agg_dens <- get_fia()[,colnames(pls_data)]

@@

Over the past 200 years centennial scale climate variability and land use change have interacted to transform the forests of the northeastern United States. There are clear indications that modern forests, as represented by the Forest and Inventory Analysis show greater homogeneity and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land use conversion in the late 1800s [@goring2015preset;@schulte2007homogenization].

This shift has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change [@iverson2013tree;@iverson1998predicting].  One of the most significant changes to forest tree distributions in the upper midwest is the loss of extensive cover through land use change [@rhemtulla2009legacies], particularly along the historical prairie and savanna margins of the region (Figure 1f), where extensive land use conversion has left little natural land.

@@

ggsave(plot = arrangeGrob(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4)), filename='figures/figure_one.tiff')

```
![figone](figures/figure_one.tiff)

@@

Methods
===============
We pair pre-settlement era vegetation data from the Public Land Survey System (PLSS) aggregated to a 64 $km^{2}$ grid [@goring2015preset] with annually resolved climate data for the region.  The PLSS era data is aggregated from survey data collected on a 1 x 1 mi grid across the Upper Midwest during the PLS, from 1830 - 1904, in a time transgressive manner, from the south east to the north west.  At each survey point surveyors noted the closest two to four trees, recorded their distances to the plot center point, diameters and the azimuth to the tree.  Goring et al. [-@goring2015preset] aggregated this data to an 8 x 8 km grid to reduce local-scale variability, resulting in a gridded data product with an average of 61 plots per cell, or approximately 120 trees per cell.  When examining relationships between FIA and PLS data, Goring et al. [-@goring2015preset] report significant relationships within the FIA relating dissimilarity from PLSS era forests and the number of FIA plots within an 8x8km cell, however this relationship accounts for only 3% of the total change in dissimilarity, indicating statistical significance, but little ecological significance on this regional scale.

@@

The analysis presented here follows a blocked sampling method. We generate probability density functions along historical climate gradients for taxa within the PLSS and for the same PLSS taxa (and point locations) using climate data for the modern era. We do the same for FIA presence (climate at the PLSS and modern eras), and for anthropogenic land use classes. Climate data is derived from monthly PRISM data from 1895 - 2014. A 30-year climate normal was generated for modern July temperature ($T_{jul}$), January temperature ($T_{jan}$), the continentality (difference between $T_{jul}$ and $T_{jan}$, $T_{diff}$), and annual precipitation ($P_{ann}$).

@@

By generating kernel density estimates of taxon presence along climate gradients drawn from each of the four states of climate and vegetation ($PLSS_{veg}$-$PLSS_{clim}$, $PLSS_{veg}$-$FIA_{clim}$, $FIA_{veg}$-$PLSS_{clim}$, $FIA_{veg}$-$FIA_{clim}$) we can use Hellinger distance to estimate the effects of changing climate ($d_{H}$), or the effect of changing land use (d_H_ between kernel densities in the same row of Figure 2).  In particular, the effect of changing land use and climate (d_H_ between the top left and bottom right in Figure 2) can be parsed as the sum of differences in the left hand column (testing climate change) and the top row (testing land use only).  The relative magnitudes of these differences gives us an indication of the influence of either land use or climate in driving shifts in species distributions along the climate gradients in the Upper Midwest.

@@

loss_table <- as.data.frame.matrix(table(loss_melt[[1]]$variable,
                                         loss_melt[[1]]$value))
loss_table <- data.frame(loss_table, 
                         Modern = round((loss_table$Gain + loss_table$Presence) * .064, 0),
                         Historical = round((loss_table$Loss + loss_table$Presence ) * .064, 0))

loss_table[,1:3] <- round(tran(loss_table[,1:3], method = 'proportion') * 100,1)

ggsave(plot = loss_melt[[2]], filename = 'figures/Figure_panelloss.tiff',
       dpi=300, width=6, height = 3)
```

![figuretwo](figures/Figure_panelloss.tiff)
**Figure 2.**.  *Gain loss and continuous presence for PLSS and FIA tree taxa. Loss is assessed as cells that contained presence during the PLSS era, and no presence in the FIA data (light blue). Continuous presence is indicated by dark gray, where both FIA and PLSS cells indicate presence.  Few taxa show novel presence (red), although taxa such as poplar, ash and basswood show gains of over 10%.*

**Table 1.** *Gain and loss of the various forest tree types of interest since the pre-settlement era. The first three columns represent percentages of the total area occupied by the taxon since settlement. Modern and Historical values represent 100s of km^2^ occupied by the tree taxon within either the FIA or PLSS data.*

@@

```{r densities, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}
source('R/resampling_climate.R')
vegclim_table <- resample_climate(climate, agg_dens, pls_data)

source('R/figures/climate_boxplots.R')
boxplot_clim <- clim_boxplots(clim_table = vegclim_table)

@@

![FigureThree](figures/boxplot_clim.tiff)
**Figure 3.** *Changes in the realized climate space for genera in the upper Midwestern United States. For forest taxa maximum temperatures decline across the region from the PLS era to the modern era.  At the same time minimum temperatures and annual precipitation have both increased.*

@@

```{r panel_densities, warning=FALSE, message=FALSE, echo=FALSE}
source('R/figures/FigureFour_Panels.R')
```

![modelgrid](figures/figureFour_panels.tiff)
**Figure 4**. *Blocked method for analysis, with representative normalized densities for climate distributions.  Data presented represent one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row).  A diagonal shift from the top left to bottom right represents the true change in forest-climate relationships.*

Hellinger distances reveal that while taxa may show the imprint of land use change on shifting climatic niches, this shift is not uniform across taxa not climatic  variables. Maximum temperature shows the greatest change attributed to land use change.  Tamarack, elm, poplar, spruce beech, cedar, and fir all show a greater influence for land use change than regional climate change. The precipitation panel (Figure 5) indicates that precipitation across the region has driven change in the species climate niche for this variable.  Only spruce and poplar show a marginally greater effect of land use change than climate change, wheras taxa such as beech and hemlock show a strong climate signal.

```{r climate_shifts, warning=FALSE, message=FALSE, echo=FALSE}

@@

```
![hellplot](figures/hellingerplot.tiff)

**Figure 5**. *For each of the four variables examined we see differences in the relative contributions of land use and climate in driving patterns of change over the settlement era.  The x-axis represents the total shift (diagonal in Figure 4) from PLSS - FIA for both vegetation and climate, while the y-axis represents the difference between $d_{h}(V_{PLSS}C_{PLSS},V_{PLSS}C_{FIA})$ and $d_{h}(V_{FIA}C_{FIA},V_{PLSS}C_{FIA})$. Taxa with negative y-axis values have changes in distributions along the climate gradient that are driven largely by land use, while positive y-axis values can be largely attributed to regional climate change.*

**Table 2**. *Compounding (+), confounding (.) and contradicting (-) relationships between land use and climate change.  There is no relationship between spatial correlates and the land use/climate interaction.*

```{r tests, warning=FALSE, message=FALSE, echo=FALSE, results='as-is'}
source('R/confounding_table.R')

pandoc.table(conf_table, justify = "left", style = 'simple')

@@

@@

Land use is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000]. Land use change reduces the correlational structure between species and climate [@devineau2011arid]. While rates of land use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land use are likely to persist, both in reorganizing forested landscapes [@goring2015preset], and in the correlational structure between species and climate. Paleoclimatic modeling indicates shifts in realized niches for species over time, potentially resulting from non-analogue climates or shifts in competition resulting from changing patterns of co-occurence through time [@maiorano2013building].

From a modern perspective, the US Forest Service's Forest Inventory and Analysis data set is an important resource for researchers, providing detailed data at fine grain over the entire contiguous United States. However, the FIA is also an incomplete record of tree species distributions in the United States since it limits plot surveys to plots with > 10% tree cover [@woudenberg2010forest].  In much the same way, climatic data plays an important role in building correlative models, but we are limited to the modern era [@pellatt2012using;@iverson1998predicting], while evidence strongly suggests historical datasets add considerably to our understanding of the breadth of species' climatic niches [@ivory2015; @maiorano2013building;@early2014climatic].
